---
title: "Table 1: Static Output Drip {{< fa droplet >}}"
format: html
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

# Setup -----------------------------------------------------------------

# Turn off scientific notation
options(scipen = 999)

# Load libraries
libraries <- c("dplyr", "readr", "ggplot2", "gt", "openxlsx", "janitor")
invisible(sapply(libraries, library, character.only = TRUE))

# Functions
program_list <- list.files("R", full.names = TRUE)
invisible(sapply(program_list, source))

# Make sure our folders exist
dir.create("data-raw", showWarnings = FALSE)
dir.create("outputs", showWarnings = FALSE)
```

Hey, analyst! Welcome to 🌧️ **Raining Tables** 🌧️, a precipitation-themed package that puddle-jumps through 
a minimal data pipeline. 

In this package, we'll:  

- read in a spreadsheet (either this package's demo file or the file you downloaded in Step 00), 
- explore various ways to output it (Excel, PDF), 
- pull data updates from the public NOAA API, 
- read out to a couple database types, 
- test our code and ensure it runs identically on other people's computers, 
- save our outputs to Google Drive, 
- create a simple dashboard, 
- put our code in a container, and 
- run it from ☁️ _the cloud._ ☁️

Today, we're taking the first small step: read in `precipitation.csv` from `data-raw/`, 
plot it in a nice chart, and export it as a table. 

One folder structure note:

Our spreadsheet lives in `data-raw/`, not mixed in with our code (`R/`) or our file outputs (`outputs/`). That's deliberate: raw data is a read-only input, never
something a script overwrites. Folder structure helps us keep our workspace clean and tidy.

Organizing our files this way makes updating a breeze ~ 🪁

## Read the raw data

```{r read_data}
rainfall_raw <- readr::read_csv("data-raw/precipitation.csv")

names(rainfall_raw)
```

## Clean it up

NOAA's spreadsheet gives us uppercase field names. Our
`plot_rainfall()` function expects lowercase columns --
specifically `date`, `prcp`, and `station`. Our first task is to 
tidy the dataset up a bit by converting all field names to standard
lowercase.

`janitor::clean_names()` does the lowercase conversion for us:

```{r clean_data}
rainfall <- rainfall_raw |>
  janitor::clean_names()

names(rainfall)
```

Field names before: `r names(rainfall_raw)[c(1, 6, 11)]`.  
Field names after: `r names(rainfall)[c(1, 6, 11)]`. 

In this case `clean_names()` alone was enough to match `plot_rainfall()`'s expected
`station`, `date`, `prcp` columns.

## Plotting raindrops 🫧

First, we plot precipitation (see `R/plot_rainfall.R` for plot code):

```{r plot_rainfall}
rainfall_plot <- plot_rainfall(rainfall)
rainfall_plot
```

Then we save out the plot as a .png in `outputs/rainfall_chart.png`:

```{r save_plot}
ggplot2::ggsave(
  filename = "outputs/rainfall_chart.png",
  plot = rainfall_plot,
  width = 8,
  height = 5
)
```

## Exporting static files

Once we update, we often have to get our data out to a wider audience, who
may only be interested in static files, like .csvs, PDFs, or Excel. We'll also save
these out to `outputs/` (see `R/export_outputs.R` for code). 

### Here's how we do that:

**A csv** is our universal, pipeline-friendly format, easy for another
script, language, or database loader to pick up:

```{r export_csv}
export_csv(rainfall, "outputs/rainfall.csv")
```

**Excel** is for handoffs to stakeholders who prefer handling Excel files.
For this, we use the openxlsx package to read out to Excel, rather than a non-proprietary csv file format.

While they're common outputs, Excel files are uncommon in data pipelines. Multiple sheets, too much styling,
embedded formulas all make this format less usable than a simple csv.

```{r export_xlsx}
export_xlsx(rainfall, "outputs/rainfall.xlsx")
```

**A gt table as a PDF** gives you a static PDF report, good for presentations and attachments.

PDFs are also common outputs, but you wouldn't use them in a data pipeline unless your files
were handed to you this way and you had no other choice. They were never designed to be
machine-readable; it's easy to misread or misinterpret PDF information. 

```{r export_pdf}
export_gt_pdf(
  head(rainfall, 15),
  "outputs/rainfall_table.pdf",
  title = "Daily Precipitation"
)
```

## What we've just output

Starting from a csv sitting in `data-raw/`, we cleaned its column names,
plotted daily precipitation, and exported the results into `outputs/`. 

Next up: rather than reading in static files, we'll pull our rain data directly from NOAA's API. ⚡ 

